---
title: "Plotting US spatial data with Alaska and Hawaii offset"
format: html
editor: visual
---

## Plotting US states with Alaska and Hawaii

When we try and make maps that include Alaska, Hawaii, or other territories outside the lower 48 states, it creates some problems for visualization (Data from ESRI [ArcGIS Hub](https://hub.arcgis.com/datasets/TrainingServices::us-state-boundaries/explore?location=24.148743%2C-65.721188%2C3.01)).

```{r}
#| message: FALSE
#| warning: FALSE 
library(tidyverse)
library(sf)

#read in spatial data
allStates<-st_read(("US_State_Boundaries.shp"))

#plot it
ggplot()+
  geom_sf(data=allStates)
```

What we may want to do is shift the geometry of these states to locations in closer proximity to the lower 48.

## Shifting geometry with `tigris` package

We can accomplish this with the `tigris` package. Here, we'll do it for the 50 US states.

```{r}
#| message: FALSE
#| warning: FALSE 
#install.packages(tigris)
library(tigris)

#filter to create a new dataset with 50 States (#26-27 are USVI and PR)
fiftyStates<-allStates %>% 
  filter(OBJECTID %in% c(1:25,28:60)) 

#Offset AK and HI using shift_geometry function from tigris
fiftyStatesShifted<-shift_geometry(fiftyStates,position=c("below"))

#plot new map
ggplot()+
  geom_sf(data=fiftyStatesShifted)

```
