---
title: "Web scraping from screenshots?"
format: html
editor: visual
---

## Packages!

We'll need these packages for this operation:

```{r}
#| warning: FALSE
#| message: FALSE
#install.packages("webshot2")
#install.packages("magick")
#install.packages("tesseract")
#install.packages("tidyverse")
library(webshot2) #for snapshots
library(magick) #for image reading
library(tesseract) #for ocr
library(tidyverse) #for data
```

## Taking screenshots of websites

The **`webshot2`** package can be used to take a screenshot of a webpage, it requires Google Chrome to be installed on the machine. The `webshot` function needs a url, add a delay to give the site time to load, and a filename to store the screenshot. Let's do it for my Tufts profile.

```{r}

webshot("https://facultyprofiles.tufts.edu/benjamin-davies", delay = 0.5,file="webshot.png")
```

## OCR website

The **`magick`** and **`tesseract`** packages can be used to read in an image, then recognize and pull text from an image.

```{r}
#read an image
img <- image_read("webshot.png")

#get the text, tell tesseract it's in English
text<-image_ocr(img,language="eng")
```

Now we can look at the text...

```{r}
text
```

It's a single giant character object with all the text on the screen. Let's say I wanted the university where the prof got their PhD. Here you have to look for patterns that will let you isolate the interesting piece of information. Blocks of text are split by `\n`; university name follows `PhD` heading, and then is followed by a comma. \

```{r}
#split up text based on \n
textSplit<-strsplit(text, "\n")
#this produces a list object, unlist turns it into a vector
textSplit<-unlist(textSplit)
#which tells me where the word "PhD" comes up in the vector
phd<-which(str_detect(textSplit,"PhD"))
#pull out the next item in the vector
school<-textSplit[phd+1]
#This splits up that object based on commas, which produces a list object
school<-str_split(school, ",")
#Make list object a vector
school<-unlist(school)
#Take the first item in the vector
school<-school[1]
#print it out
school
```

## Trying a scrape

Now we're going to try and get alma maters for the ENVS faculty. We'll create a vector that includes the url ID for each faculty member, then use a `for` loop to iterate through that vector and run the code above for each profile.

```{r}
#list of profs in alpha order
profs<-c("benjamin-davies","sara-gomez","hassaanfurqan-khan","karen-kosinski","ninian-stein")
#empty vector to store schools
almaMater<-c()

#This loop iterates over the profs vector, pasting their name into the url to get their profile, then runs the code above (though see below where tail and append function used). After this, the result gets added to a vector of schools.
for (i in profs){
  url<-paste0("https://facultyprofiles.tufts.edu/",i)
  webshot(url, delay = 0.5,file="webshot.png")
  img <- image_read("webshot.png")
  text<-image_ocr(img,language="eng")
  data<-image_ocr_data(img,language="eng")
  textSplit<-strsplit(text, "\n")
  textSplit<-unlist(textSplit)
  phd<-which(str_detect(textSplit,"PhD"))
  school<-textSplit[phd+1]
  #This line was added because some folks mentioned the phrase PhD in their bio, so tail just finds the last instance in the text
  school<-tail(school,n=1)
  school<-str_split(school, ",")
  school<-unlist(school)
  school<-school[1]
  #This adds the result to the almaMater vector
  almaMater<-append(almaMater,school)
}


```

Now combine the profs and schools into a tibble, and voilà!

```{r}
#remove the hyphens in prof names
profs<-str_replace(profs,"-"," ")
#capitalize first letters
profs<-str_to_title(profs)
#make tibble
profPhD<-tibble(profs,almaMater)
profPhD
```
